---
title: "Emily Analysis"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: yes
    toc_depth: 5
    toc_float: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)

library(readxl)
library(tidyverse)
library(knitr)
library(kableExtra)
library(table1)
library(janitor)
library(patchwork)
library(gtsummary)

options(
  ggplot2.continuous.colour = "viridis",
  ggplot2.continuous.fill = "viridis"
)

scale_colour_discrete = scale_colour_viridis_d
scale_fill_discrete = scale_fill_viridis_d

options(scipen = 999)
```

TO DO:

EDA:
- outcome variables: 

	whether the medications are safe for patients
	
	whether patients could easily adhere to medication schedules so that long term use is feasible.
	
	assess the correlation of PK with adherence measures and the occurrence of adverse events 
	
	demographic factors associated with product adherence and whether they differ by product used (Pill or gel) or regimen (three times a day or once a day)
	
- spaghetti plot of weeks —> linear or non-linear

Mixed Model(s):
- sequence/carry-over??

Contrasts

```{r}
baseline <- read.csv("baseline.csv") %>%
  dplyr::select(-X) %>%
  mutate(carryover_period2 = period1,
         carryover_period3 = period2)


baseline_long <- baseline %>%
  pivot_longer(cols = period1:period3,
               names_to = "period",
               values_to = "treatment") %>%
  mutate(bloodVL_before = case_when(
            period == "period1" ~ bviral0,
            period == "period2" ~ bviral2,
            period == "period3" ~ bviral4),
         bloodVL_after = case_when(
            period == "period1" ~ bviral1,
            period == "period2" ~ bviral3,
            period == "period3" ~ bviral5),
         skinVL_before = case_when(
            period == "period1" ~ sviral0,
            period == "period2" ~ sviral2,
            period == "period3" ~ sviral4),
         skinVL_after = case_when(
            period == "period1" ~ sviral1,
            period == "period2" ~ sviral3,
            period == "period3" ~ sviral5),
         carryover = case_when(
            period == "period1" ~ "0",
            period == "period2" ~ carryover_period2,
            period == "period3" ~ carryover_period3)) %>%
  dplyr::select(-c(bviral0:carryover_period3)) %>%
  pivot_longer(cols = c(bloodVL_before:skinVL_after),
                          names_to = c(".value", "timepoint"),
                          names_sep = "_") %>%
  mutate(timepoint = ifelse(timepoint == "before", 0, 4))

write.csv(baseline_long, "baseline_longformat.csv")
```

```{r}
ggplot(baseline_long) +
  geom_line(aes(x = timepoint, y = bloodVL, group = ptid, color = period)) +
  facet_wrap(~ treatment) +
  theme_bw()


baseline %>%
  dplyr::select(age:gender) %>%
  tbl_summary(label = list(
      age ~ "Age (years)",
      race ~ "Race",
      gender ~ "Female"))
```


Preliminary Mixed Model - Issues with Period & Carryover (how to rewrite?)

PK ~ Random Intercept (Subject) + Treatment*Timepoint + Carryover + Period

```{r}
library(nlme)

fit2 <- lme(bloodVL ~ treatment * timepoint + carryover, method="ML",
           data = baseline_long,
          random=reStruct(~ 1 | ptid))

summary(fit2)
```


```{r}
endpoints <- read.csv("endpoints.csv") %>%
  mutate(carryover_period2 = period1,
         carryover_period3 = period2) 

endpoints_long <- endpoints %>%
  pivot_longer(cols = period1:period3,
               names_to = "period",
               values_to = "treatment") %>%
  mutate(
    AE_pillA_total = rowSums(select(., starts_with("AE_pillA")), na.rm = TRUE),
    AE_gelB_total = rowSums(select(., starts_with("AE_gelB")), na.rm = TRUE),
    AE_gelC_total = rowSums(select(., starts_with("AE_gelC")), na.rm = TRUE),
    overall_safety = case_when(
      treatment == "Pill A" ~ factor(ifelse(AE_pillA_total > 0, 1, 0)),
      treatment == "Gel B" ~ factor(ifelse(AE_gelB_total > 0, 1, 0)),
      treatment == "Gel C" ~ factor(ifelse(AE_gelC_total > 0, 1, 0))),
    adhere_week1 = case_when(
      treatment == "Pill A" ~ Adhere_pillA_week1,
      treatment == "Gel B" ~ Adhere_gelB_week1,
      treatment == "Gel C" ~ Adhere_gelC_week1),
    adhere_week2 = case_when(
      treatment == "Pill A" ~ Adhere_pillA_week2,
      treatment == "Gel B" ~ Adhere_gelB_week2,
      treatment == "Gel C" ~ Adhere_gelC_week2),
    adhere_week3 = case_when(
      treatment == "Pill A" ~ Adhere_pillA_week3,
      treatment == "Gel B" ~ Adhere_gelB_week3,
      treatment == "Gel C" ~ Adhere_gelC_week3),
    adhere_week4 = case_when(
      treatment == "Pill A" ~ Adhere_pillA_week4,
      treatment == "Gel B" ~ Adhere_gelB_week4,
      treatment == "Gel C" ~ Adhere_gelC_week4),
    carryover = factor(case_when(
      period == "period1" ~ "0",
      period == "period2" ~ carryover_period2,
      period == "period3" ~ carryover_period3))) %>%
  dplyr::select(-c(AE_pillA_week1:carryover_period3), -c(AE_pillA_total:AE_gelC_total)) %>%
  pivot_longer(cols = starts_with("adhere"),
               names_to = "timepoint",
               values_to = "adherence") %>%
  mutate(timepoint = as.numeric(str_extract(timepoint, "\\d+")))


write.csv(endpoints_long, "endpoints_longformat.csv")

```
